A weighted sampling algorithm for the design of RNA sequences with targeted secondary structure and nucleotides distribution

### 2013-03-12 ISMB/results_ISMB.tex
intermediate commit 2

\subsection{Success rate}
We started by estimating the success rate of our methodology and computed the percentage of sequences with a MFE structure identical to the target secondary structure. Figure~\ref{fig:mfe_struct_solved_noinverse} shows our results. We clearly see that before the post-processing step (i.e. \RNAinverse) the sequences sampled by \ourprog have a low success rate (first row). As mentioned earlier, this could be explained by the fact that no selection criterion has been at this stage applied to unpaired nucleotides. Remarkably, after the local search optimization (with \RNAinverse) of nucleotides in unpaired regions (second row), we observe a dramatic improvement of our success rate. As expected, we observed that length is, in general, not a good predictor for the hardness of designing a structure. Instead, a high number of free nucleotides in the structure seems to be a  good measure of the hardness of its design. Similarly, these data also show that designing sequences with low \GCContent is challenging for all types of targets.
}

@@

\highlight{
Using the same dataset of $50$ structures, we generated $100$ samples per structure with classical softwares who do not control the \GCContent. Namely,  \RNAinverse, \INFORNA, \NUPACK and \frankenstein.  We show the distribution of the \GCContent of the sequences produced with these softwares in Fig.~\ref{fig:gcdist} those distributions. As anticipated, we observe a clear bias toward high {\GCContent}s and a complete absence of sequence with less than $30\%$ of \GC. This striking results motivates a need for methods that enable to explicitly control the \GCContent and more precisely that enable to design sequences with low \GCContent (i.e. 30\% or less).
}

\highlight{
A desirable feature in sequence design, is to produce samples with a high sequence diversity and stable secondary structure. 
Therefore, in the following we will use two useful measures which are the sequence identity of the samples, and the Boltzmann probability
of the target structure in the low energy ensemble.
}
%The positional entropy is defined as:
%\begin{align}
%  \sum_{\substack{i\\b \in (\Ab,\Cb,\Gb,\Ub)}} -p_{i\to b}\cdot \log_4(p_{i\to b}) &&& \text{(Positional entropy)}
%  \label{eq:entropy}
%\end{align}
%where $p_{i\to b}$ is the proportion of nucleotides $b$ at position $i$.
%Intuitively, this measure captures the variability of sequences generated by a given method.
\highlight{
The sequence identity is defined over a set $\mathcal{S}$ of aligned sequences (in our case, all sequences have the same length and can be trivially aligned) as :
\begin{align}
  \sum_{s^1,s^2 \in\mathcal{S}\times\mathcal{S}}\left(\frac{1}{|s^1|}{\sum_{\substack{i\\s^1_i\equiv s^2_i}}1}\right) &&& \text{Seq. identity}
\end{align}
where $s_i$ is the nucleotide at position $i$ in sequence $s$. Intuitively, this measure captures the diversity of sequences generated by a given method. Next, the Boltzmann frequency is defined, for a structure $\Struct$ and a sequence $s$ as:
\begin{align}
  e^{\frac{-\ES(s,\Struct)}{RT}}/\mathcal{Z}^s  &&&\text{Frequency}
\end{align}
where $\mathcal{Z}^s$ is the partition function of sequence $s$. This measure tells us how dominant is a structure $\Struct$ in the Boltzmann ensemble of structure over a sequence $s$. A high value implies a stable structure. We compute this frequency with \rnafold from the \texttt{Vienna Package 2.0} \citep{Hofacker:1994}.
}

@@

  Figure~\ref{fig:nb_sols_entropy} shows the number of correct
solutions as their average entropy and positions inside base pairs entropy, since 
\ourprog  only influences the distribution of nucleotides inside 
stacked base pairs. The most constraining factor for generating valid
 solutions seems to be  low \GCContent and the percentage of free nucleotides.

%Our method is able to generate

Also, a critical property in RNA sequence design is 
the frequency of the MFE. 
The sequence should be in its target conformation long enough to
perform the desired action. Presented in Figure~\ref{fig:freq} we see that
there is a slow decline of the frequency with the increase in size. Yet,
for an average \GCContent, the frequency stays over $10\%$ even
at size of $100$ nucleotides.

\begin{figure*}[ht!]
	\centering
	\includegraphics[width=\textwidth]{Figures/nb_sols_entropy}
	\caption{Number of solutions generated and their average entropy. 
	The last row presents only the entropy of the nucleotides inside base 
	pairs.}
	\label{fig:nb_sols_entropy}
\end{figure*}

\begin{figure*}[ht!]
	\centering
	\includegraphics[width=\textwidth]{Figures/freq.png}
	\caption{Average Boltzmann probability of the MFE.}
	\label{fig:freq}
\end{figure*}

@@

\subsection{Global sampling vs Local search vs Glocal approach}

Here, we are interested in estimating the impact of the design methodology on the performances. More precisely, we aim to determine the merits of a global sampling approach (\ourprog), compared to a glocal procedure (\ourprog + \RNAinverse) and a local search methodology (\RNAinverse).

It should be noted, however, than \RNAinverse produces sequences whose \GCContent is extremely concentrated (see Fig.), thereby requiring a large number of runs before producing a suitable \GCContent sequence. Therefore, a rejection algorithm based on \RNAinverse alone should not be considered a realistic solution for the design with constrained \GCContent.

{\em Ajouter considérations Base Pair entropy: Se rappeler, que moins, c'est mieux !}

>>>>

Before starting our benchmark, we asses the need for our methods and performed an analysis of the \GCContent drift achieved with state-of-the-art software. Using our dataset of $50$ structures, we generated $100$ samples per structure with classical softwares who do not control the \GCContent. Namely,  \RNAinverse, \INFORNA, \NUPACK and \frankenstein.  We show the distribution of the \GCContent of the sequences produced with these softwares in Fig.~\ref{fig:gcdist} those distributions. As anticipated, we observe a clear bias toward high {\GCContent}s and a complete absence of sequence with less than $30\%$ of \GC. This striking results motivates a need for methods that enable to explicitly control the \GCContent and more precisely that enable to design sequences with low \GCContent (i.e. 30\% or less).
}

\highlight{
\subsection{Success rate}
We started by estimating the success rate of our methodology and computed the percentage of sequences with a MFE structure identical to the target secondary structure. Figure~\ref{fig:mfe_struct_solved_noinverse} shows our results. We clearly see that before the post-processing step (i.e. \RNAinverse) the sequences sampled by \ourprog have a low success rate (first row). As mentioned earlier, this could be explained by the fact that no selection criterion has been at this stage applied to unpaired nucleotides. Remarkably, after the local search optimization (with \RNAinverse) of nucleotides in unpaired regions (second row), we observe a dramatic improvement of our success rate. As expected, we observed that length is, in general, not a good predictor for the hardness of designing a structure. Instead, a high number of free nucleotides in the structure seems to be a  good measure of the hardness of its design. Similarly, these data also show that designing sequences with low \GCContent is challenging for all types of targets.
}

@@

%The positional entropy is defined as:
%\begin{align}
%  \sum_{\substack{i\\b \in (\Ab,\Cb,\Gb,\Ub)}} -p_{i\to b}\cdot \log_4(p_{i\to b}) &&& \text{(Positional entropy)}
%  \label{eq:entropy}
%\end{align}
%where $p_{i\to b}$ is the proportion of nucleotides $b$ at position $i$.
%Intuitively, this measure captures the variability of sequences generated by a given method.

\highlight{
A desirable feature in sequence design, is to produce samples with a high sequence diversity and stable secondary structure. 
Therefore, in the following we will use two useful measures which are the sequence identity of the samples, and the Boltzmann probability
of the target structure in the low energy ensemble.

The sequence identity is defined over a set $\mathcal{S}$ of aligned sequences (in our case, all sequences have the same length and can be trivially aligned) as :
\begin{align}
  \sum_{s^1,s^2 \in\mathcal{S}\times\mathcal{S}}\left(\frac{1}{|s^1|}{\sum_{\substack{i\\s^1_i\equiv s^2_i}}1}\right) &&& \text{Seq. identity}
\end{align}
where $s_i$ is the nucleotide at position $i$ in sequence $s$. Intuitively, this measure captures the diversity of sequences generated by a given method. Next, the Boltzmann frequency is defined, for a structure $\Struct$ and a sequence $s$ as:
\begin{align}
  e^{\frac{-\ES(s,\Struct)}{RT}}/\mathcal{Z}^s  &&&\text{Frequency}
\end{align}
where $\mathcal{Z}^s$ is the partition function of sequence $s$. This measure tells us how dominant is a structure $\Struct$ in the Boltzmann ensemble of structure over a sequence $s$. A high value implies a stable structure. We compute this frequency with \rnafold from the \texttt{Vienna Package 2.0} \citep{Hofacker:1994}.
}

@@

%Figure~\ref{fig:nb_sols_entropy} shows the number of correct solutions as their average entropy and positions inside base pairs entropy, since \ourprog  only influences the distribution of nucleotides inside stacked base pairs. The most constraining factor for generating valid solutions seems to be  low \GCContent and the percentage of free nucleotides.

\highlight{
Figure~\ref{fig:nb_sols_entropy} shows the number of solutions generated (i.e. sequences with a MFE structure identical to the target structure) and the average sequence identity of the sequences. Here, we note that low {\GCContent}s have a strong (negative) influence on the number of sequences generated, and in parallel also affect negatively the sequence diversity. This observation emphasizes the difficulty to  design sequences with low \GCContent. Once again, large percentages of free nucleotides increase the difficulty of the task.

The thermodynamical stability of the target structure on the designed sequence is another important property when estimating the performance of RNA design algorithms. We estimate the quality of our solutions in Figure~\ref{fig:freq}. First, we observe a slow decline of the structure stability (i.e. the frequency) when the target structure increases in size. Yet, for an average \GCContent, the frequency stays over $10\%$ even at size of $100$ nucleotides. Next, we note that for the most difficult target structures (i.e. the longer ones or those with high percentages of unpaired nucleotides in loops) the \GCContent have a limited (almost null) influence on the stability of the target structure on the designed sequence. By contrast, this less true for easiest and small structures with only few free nucleotides in internal loops.
}

\begin{figure*}[ht!]
	\centering
	\includegraphics[width=\textwidth]{Figures/nb_sols_entropy}
	\caption{Number of solutions generated by \ourprog and their average sequence identity. The curves are parameterized by the {\GCContent} of the sequences. The data are plotted vs. the length of the target (left), density of stacked base pairs (centre) and number of unpaired nucleotides  in loops (right).}
	\label{fig:nb_sols_entropy}
\end{figure*}

\begin{figure*}[ht!]
	\centering
	\includegraphics[width=\textwidth]{Figures/freq.png}
	\caption{Thermodynamical stability of the target structure. The curves report the average Boltzmann probability of the target structure (which is also the MFE structure) at various {\GCContent}s w.r.t. the length of the target (left), density of stacked base pairs (centre) and number of unpaired nucleotides  in loops (right).}
	\label{fig:freq}
\end{figure*}

\highlight{
\subsection{Global sampling vs Local search vs Glocal approach}
To conclude this study, we estimate the impact of the design methodology on the performances. More precisely, we aim to determine the merits of a global sampling approach (\ourprog), compared to a glocal procedure (\ourprog + \RNAinverse) and a local search methodology (\RNASSD). To our knowledge, \RNASSD is the only other software that implements an explicit control of the \GCContent. We focus our analysis on the design of sequences with low {\GCContent}s (30\% and less) as they are almost impossible to design with classical software (See Figure~\ref{fig:gcdist}).

It should be noted, however, than \RNAinverse produces sequences whose \GCContent is extremely concentrated (see Fig.), thereby requiring a large number of runs before producing a suitable \GCContent sequence. Therefore, a rejection algorithm based on \RNAinverse alone should not be considered a realistic solution for the design with constrained \GCContent. 
}

@@
